Trait plasticity and covariance along a continuous soil moisture gradient

### 2021-01-17 FVT Manuscript/manuscript.Rmd
V5

author: 
  - name          : "J. Grey Monroe"
    affiliation   : "1,4"
    corresponding : yes    # Define only one corresponding author
    address       : "Current address: 1 Max Planck Ring, 72076 Tubingen, DE"
    email         : "greymonroe@gmail.com"
  - name          : "Haoran Cai"
    affiliation   : "2"
  - name          : "David L. Des Marais"
    affiliation   : "2,3"
affiliation:
  - id            : "1"
    institution   : "Department of Plant Sciences, University of California at Davis, Davis, USA"
  - id            : "2"
    institution   : "Department of Civil and Environmental Engineering, Massachusetts Institute of Technology, Cambridge, MA, USA"
  - id            : "3"
    institution   : "The Arnold Arboretum of Harvard University. Boston, MA, USA"
  - id            : "4"
    institution   : "Max Planck Institute for Developmental Biology, Tubingen, DE"

abstract: Water is perhaps the greatest environmental determinant of plant yield and fitness, yet our understanding of plant-water relations is limited because it is primarily informed by  experiments that have treated soil moisture variability  as two simple discrete levels - wet and dry - rather than a continually varying environmental gradient. Here we used experimental and statistical methods based on function-valued traits to study responses to continuously varying soil moisture gradient in a suite of physiological and morphological traits in two species and multiple genotypes of the model grass \textit{Brachypodium}. We find that the majority of traits exhibit non-linear responses to soil moisture variability. We also observe differences in the shape of these non-linear responses between traits, species, and genotypes. Emergent phenomena arise from this variation including changes in trait correlations and evolutionary constraints as a function of soil moisture. These results point to the importance of considering non-linearity in plant-water relations to predict plastic and evolutionary responses to changing climates.

@@

For plants, soil water availability is one of the most important environmental factors in ecology and agriculture, acting as a major determinant of fitness and yield [@juenger2013natural; @greenham2017temporal]. Considerable interest and effort have been placed on studying plant responses to drought, often defined conceptually and experimentally as an environmental condition of abnormally elevated aridity resulting in decreased plant performance [@passioura1996drought]. Most of the vast research on drought responses and tolerance strategies, including in _Brachypodium_, the focal system of this work, has been limited to comparisons between discrete soil water levels - control and water-limited [@des2012physiological; @edwards2012quantitative; @el2014genotype; @greenham2017temporal; @vasseur2014multivariate]. Yet soil moisture as an environmental factor is complex and multidimensional, with fluctuations varying continuously in timing, duration, and degree. Here we investigate trait responses to one important dimension of soil moisture variability - degree - with experimental and statistical approaches treating soil moisture content as a continuous variable rather than a set of fixed levels.

The limited extent to which continuous environmental variables have been studied is not unique to investigations of plant responses to soil moisture. Recent decades have seen increased appreciation for modeling traits as mathematical functions of continuous variables across diverse organisms [@rocha2012connecting; @pettay2008age; @mason2020learning], a wide range of traits [@robinson2009impact; @stinchcombe2010across; @mcguigan2010quantitative], and numerous components of the environment [@brommer2008exploring; @mcguigan2009condition; @pearse2019life]. This burgeoning interest in the importance of "function-valued traits" has evolved from a call-to-arms to a field experiencing rapid theoretical and empirical advances [@kirkpatrick1989quantitative; @kingsolver2001variation; @griswold2008hypothesis; @stinchcombe2012genetics; @goolsby2015phylogenetic; @gomulkiewicz2018variation]. For plant biologists this presents a compelling framework for gaining a deeper understanding of plant adaptations to the environment.

A deep body of literature aims to understand the traits and strategies used by plants to acclimate to variable soil water content. Quantitative genetic variation in myriad drought resistance traits have been observed in natural populations and laboratory model systems. In particular, natural populations of Brassicaceae species (including _Arabidopsis thaliana_) and _Brachypodium_ harbor variation both constitutive and inducible traits mediating plant-water relations, including water use efficiency [@des2012physiological; @edwards2012quantitative; @greenham2017temporal; @des2017interactive], leaf chemistry [@des2017interactive; @kesari2012intron], leaf anatomy [@skirycz2011pause; @verelst2013molecular;@dittberner2018natural], root-shoot biomass partitioning [@des2012physiological; @des2017interactive], and many others [@verslues2011drought; @edwards2012quantitative; @juenger2013natural; @yarkhunova2016selection; @luo2016specific; @lenk2019transcriptional]. Among the numerous plant traits that can and have been measured, few have been studied more extensively than specific leaf area (SLA - often reported as it's inverse, leaf mass per area or LMA). SLA provides a description of leaf architecture that is central to the leaf economics spectrum, a theory which seeks to explain variation in leaf physiological strategies, from more conservative (low SLA) to more productive (high SLA) [@wright2004worldwide]. In the context of drought stress, it is intuitive that low SLA might be adaptive as lower leaf surface area is expected to reduce water loss through transpiration. In several cases, reductions in SLA have been reported under drought conditions [@casper2001drought]. For all of these traits, the shape of plastic responses to variation in water availability remain largely unmeasured.

Despite abundant evidence of natural variation in organism-environment interaction, little is known as to how trait covariance changes across continuous environmental gradients [but see @robinson2009impact]. Describing this structure is important as trait variances and covariances can result in differential evolutionary constraints in alternative environments. For example, if total genetic variance in trait space changes depending on the environment, then the capacity to respond to selection will vary accordingly, with reduced responses to selection under conditions where trait variation is lower and vice versa. Similarly, if trait covariances depend on environment, conditions which increase trait covariances may limit evolutionary potential across a range of environments by reducing the effective axes of variation [@gomulkiewicz2018variation; @kingsolver2003environmental; @levins1968evolution; @via1985genotype]. These phenomena are made more complex by the possibility that the relationship between trait variances and covariances with the environment may be non-linear. Investigating complexities of genetic architecture is therefore needed for developing more complete quantitative predictions of evolvability in rapidly changing environments. Fortunately, advances have been made in statistical methods to reduce the dimensionality of genetic variance covariance matrices and produce meaningful summaries describing evolutionary constraints [@houle1992comparing; @blows2007tale; @kirkpatrick2009patterns; @kingsolver2015genetic].

In this study we combined these approaches with function-valued trait analyses to investigate trait plasticity, covariance and evolutionary constraints across soil moisture gradients. We model trait responses to a continuous gradient of soil moisture for a range of genotypes of two species of the model grass genus _Brachypodium_. Because the shapes of trait responses cannot necessarily be known _a priori_, we use model selection among linear and non-linear environmental predictors to estimate the response function for each trait. We then estimate genotype means at different levels of soil moisture and compute the variance-covariance parameters for all traits. Finally, we ask whether patterns of variance and covariance in drying-responsive traits in _Brachypodium_ species may lead to variation in evolutionary constraint as a function of soil moisture.

@@

_Brachypodium_ is a model genus for the genetics and genomics of C3 grasses [@brkljacic2011brachypodium]. In this work we studied natural variation between and among two species of _Brachypodium_: the annual _B. distachyon_ and the perennial _B. sylvaticum_. Both species are endemic to Eurasia, with _B.distachyon_ more prevalent in seasonally dry habitats in Southern Europe, North Africa, and the Middle East, and _B. sylvaticum_ more widely distributed throughout Eurasia [@catalan2016phylogeny] (Figure \@ref(fig:maps)).

Five genotypes of each species were studied to characterize patterns of variation in plant traits across an environmental gradient: _Brachypodium distachyon_ inbred lines ABR2, Adi-10, Bd21, Bd3-1, and Koz-1 and _Brachypodium sylvaticum_ inbred lines Ain-1, Ast-1, Kry-1, Osl-1, Vel-1. For each species, these genotypes represent a range of geographical origins and phenotypic diversity [@steinwand2013sylvaticum; @des2017interactive]. Both species are self-compatible and each of the lines used here have been maintained as inbred lines for greater than six generations [@steinwand2013sylvaticum; @vogel2009development]; as such, experimental replicates may be considered nearly homozygous.

@@

Dry down treatments began 29 and 42 days after sowing (DAS) for _B. distachyon_ and _B. sylvaticum_, respectively. Because the harvesting was divided over five consecutive days (see section below), plants were split into five equal harvest cohorts, with each cohort containing equal numbers of each watering treatment to avoid confounding harvest day with soil moisture content. Thus, though each consecutive cohort differed in age by a single day, each experienced the dry down treatment for the same amount of time. Nevertheless, we expected the difference in age between harvest cohorts to potentially impact trait expression and we therefore included harvest day (cohort) as a covariate in subsequent models. To generate a continuous gradient of final soil moisture by the end of the dry down period, plants were split into five watering treatments, receiving approximately 0, 4, 8, 12, 16, or 20 ml of water per day for 14 days. Prior to initiating the experiment pots were weighed with dry soil ($mass_{dry}$) and field capacity ($mass_{max}$) at the beginning of the experiment. These measurements provide the basis for calculating the gravimetric soil moisture content on each day and at the conclusion of the experiment. During the course of the dry down experiment, soil moisture content was calculated during the morning of day $d$ for each pot as $mass_d/(mass_{max}-mass_{dry})$.

@@

To characterize phenotypic responses to our experimental soil moisture gradient, we measured a suite of developmental and physiological traits. Plants were harvested in five cohorts over five days. Each day, half of the sampled plants were harvested for above and below ground biomass, total above ground green area, $\delta C_{13}$, N content, C content. The other half were harvested and assessed for specific leaf area (SLA) and relative water content (RWC). Above ground leaf area was estimated by laying freshly harvested plants flat between plates of clear plexiglass and imaging with a Nikon 5300 digital camera at a fixed distance with a 35mm Nikkor lens. Total green pixels were counted for each image with Easy Leaf Area (https://github.com/heaslon/Easy-Leaf-Area) with settings shown in Figure \@ref(fig:leafarea). Above ground biomass was measured after drying leaf material overnight at 60°C and then for several weeks at room temperature. Below ground biomass was measured after washing the soil matrix from roots and drying them overnight at 60°C and then for several weeks at room temperature. Above and belowground biomass was measured after leaves and roots were dried. Leaf tissues for $\delta C_{13}$, $\delta N_{15}$, nitrogen (hereafter "N") content, and carbon (hereafter "C") content were ground to a fine powder and processed by the UC Davis Stable Isotope Facility. SLA was calculated by scanning the two youngest fully emerged leaves with a 1 $cm^2$ red square. Leaf area in $mm^2$ was calculated from these same images using Easy Leaf Area. SLA was calculated as $leaf\ area / biomass_{dry}$. These leaves were also used to calculate RWC. Prior to drying, fresh leaves were weighed ($biomass_{fresh}$) and then submerged under water in 15mL falcon tubes for several hours. They were then weighed ($biomass_{turgid})$, oven-dried overnight, and weighed again ($biomass_{dry}$). RWC was calculated as $(biomass_{fresh}-biomass_{dry})/(biomass_{turgid}-biomass_{dry})$.

@@

pretty_table <- capture.output(apa_table(
 t_model_selection_summaries_byspecies
 , caption = "Model selection. *** = predictor variable p < 1e-5, ** = predictor variable p < 1e-3, * = predictor variable p < 0.5, - = predictor variable included in selected model but p > 0.05. H = harvest day, S = species, G = genotype, E = final soil moisture, E^2 = quadratic parameter, ns(E) = 2nd degree natural spline parameter",
 digits=3,
 font_size="scriptsize",
 longtable = F,
 landscape=F,
 align="c",
 midrules=1:nrow(t_model_selection_summaries_byspecies)
))

@@

### Function-valued traits
For the purposes of modeling phenotypic responses to variation in soil moisture content, we considered soil moisture content as the final soil moisture on day 14 of the dry down period for each plant, referred to in figures as $Soil\ moisture (\%_{final})$. A major challenge in studying function-valued traits is model selection. That is, identifying the functions that best describe the curvature (or lack thereof) in the shape of phenotypic responses to environmental gradients. Quadratic and natural splines have been suggested as potential functions to model non-linearities [@meyer2005randomregression], but assuming the appropriate function is problematic. Akaike information criterion (AIC) selection based on contrasting multiple complex models offers an effective means to balance predictability with over-fitting [@griswold2008hypothesis; @gomulkiewicz2018variation]. Thus, we began with the complex model for traits as described below.

$Trait = H+G+E+E^2+ns(E)_{df=2}+G*E+G*E^2+G*ns(E)_{df=2}$

where $H = harvest\ day, \ G = genotype,\ E = soil\ moisture,\ E^2 = quadratic\ parameter,\ ns(E)_{df=2} = second\ degree\ natural\ spline\ parameter$

@@

### Genetic correlations as a function of soil moisture content
We calculated trait correlations at different levels of soil moisture to characterize how genetic correlations between traits vary as a function of soil moisture content. Predicted genotypic means for each trait were calculated at 20 levels of soil moisture content (from 0.3 to 1.0 gravimetric water content) based on the model chosen by AIC (see above). Next at each level of soil moisture, pairwise Pearson correlation coefficients between genotype means were calculated within each species.

### Plasticity through multidimensional trait space
We quantified total plasticity through multidimensional trait space as a function of soil moisture by scaling each trait to a mean of 0 and calculating distance matrix between genotype means at all soil moisture levels. We looked specifically at total plasticity between consecutive soil moisture levels for each genotype. At each level of soil moisture, we then compared the two species by T-tests. To visualize plasticity of each genotype through multivariate trait space further, we performed a principal component analysis from the matrix of scaled genotype trait means using the _prcomp_ function in R.

@@

We calculated several statistics summarizing evolutionary constraint as described in @kirkpatrick2009patterns. First, for each species, we estimated the $G$ matrix of genetic covariances between mean scaled genotype trait means at different levels of soil moisture. We then calculated, using the _prcomp_ function in R, the eigenvalues of each mean standardized (trait values divided by mean) $G$ matrix, $\lambda_i$. From these we then calculated the _number of effective dimensions_, $n_{D}$, equal to the sum of the eigenvalues divided by the largest eigenvalue:

$$n_{D} = \sum_{n=1}^{n} \lambda_i/\lambda_l$$

@@

$$v_T = \sum_{n=1}^{n} \lambda_i$$

@@

(ref:curves) **Variation in phenotypic responses to soil moisture gradient as function-valued traits.** _B. sylvaticum_ genotypes are colored orange and _B. distachyon_ blue. (a) `r real_trait_names[1]` (b) `r real_trait_names[2]` (c) `r real_trait_names[3]` (d) `r real_trait_names[4]` (e) `r real_trait_names[5]` (f) `r real_trait_names[6]` (g) `r real_trait_names[7]` (h) `r real_trait_names[8]` (i) `r real_trait_names[9]` (j) `r real_trait_names[10]` (k) `r real_trait_names[11]` (l) `r real_trait_names[12]`

@@

Across the six watering treatments, combined with random variation in water capacity of pots (Figure \@ref(fig:pots)), the dry down period resulted in a continuous environmental gradient of final soil moisture, but with a higher frequency of plants near the driest extreme of soil moisture variation (Figure \@ref(fig:drydown)). This gradient provides the basis for analyzing phenotypes in relation to soil moisture treated as a continuous gradient rather than limited set of discrete factors.

The observed reduction in leaf relative water content under the driest conditions in both _B. distachyon_ and _B. sylvaticum_ indicates that at this extreme, plants were physiologically stressed (Figure \@ref(fig:curves)A). Mean leaf RWC for plants in the 10% tail of soil water content was `r tailrwc`% which is drier than that observed in the dry treatment of Des Marais et al. [@des2017interactive]. Additional observations made during the experiment such as leaf rolling, another symptom of dehydration stress, was evident in plants at the lowest water treatment by the end of the dry down period.

(ref:pca) **Plasticity through multivariate trait space.** (a) Plasticity across all traits was calculated as distance between scaled phenotype for each genotype between different levels of soil moisture. Box plots indicate species median and 25th and 75th percentiles with whiskers extending to 1.5 times the interquartile range. "*" indicates significant differences between species (t-test, $\alpha < 0.05$). Principal component analysis of scaled phenotypic responses to soil moisture gradient among genotypes of both species. Upper panels show genotype means across soil mositure content. Percent values in axis titles indicate percent variance explained by that principal component. Lower panels show eigenvectors of each trait. (b) PC1 and PC2. (c) PC3 and PC4.

### Non-linearity in trait responses to soil moisture is pervasive

We evaluated the degree to which traits show linear or non-linear shapes using an AIC model selection approach from a full model which included quadratic and natural spline parameters relating soil moisture content to plant phenotypes. We observed significant ($\alpha < 0.05$) non-linear components (quadratic, spline, or both) in the final models for all of the traits which included an environmental (water content) predictor (Table \@ref(tab:models), Figure \@ref(fig:curves)) except N content in _B. distachyon_, C content in _B. sylvaticum_, and shoot mass in _B. sylvaticum_ where non-linear environmental predictors were included in the final models chosen by AIC but did not significantly explain trait variance (p > 0.05). In _B. distachyon_, all of the traits included at least one non-linear environmental predictor. In contrast, SLA, total biomass and shoot mass did not include environmental predcitor in selected models in _B. sylvaticum_. Interestingly, of all the traits which were predicted by environment in both _B. distachyon_ and _B. sylvaticum_, the shape -- whether a quadratic versus a spline function was included in the model -- differed between the two species with the exception of carbon to nitrogen ratio.

>>>>

author: 
  - name          : "J. Grey Monroe"
    affiliation   : "1,4"
    corresponding : yes    # Define only one corresponding author
    address       : "150 California Ave, Davis, CA 95616"
    email         : "gmonroe@ucdavis.edu"
  - name          : "Haoran Cai"
    affiliation   : "2"
  - name          : "David L. Des Marais"
    affiliation   : "2,3"
affiliation:
  - id            : "1"
    institution   : "Department of Plant Sciences, Univeƒrsity of California at Davis, Davis, USA"
  - id            : "2"
    institution   : "Department of Civil and Environmental Engineering, Massachusetts Institute of Technology, Cambridge, MA, USA"
  - id            : "3"
    institution   : "The Arnold Arboretum of Harvard University. Boston, MA, USA"
  - id            : "4"
    institution   : "Max Planck Institute for Developmental Biology, Tubingen, DE"

abstract: Water availality is perhaps the greatest environmental determinant of plant yield and fitness. Yet our understanding of plant-water relations is limited because it is primarily informed by experiments considering soil moisture variability at two discrete levels - wet and dry - rather than as a continually varying environmental gradient. Here we used experimental and statistical methods based on function-valued traits to explore responses to continuously varying soil moisture gradient in physiological and morphological traits in two species and five genotypes each of the model grass \textit{Brachypodium}. We find that most traits exhibit non-linear responses to soil moisture variability. We also observe differences in the shape of these non-linear responses between traits, species, and genotypes. Emergent phenomena arise from this variation including changes in trait correlations and evolutionary constraints as a function of soil moisture. These results point to the importance of considering non-linearity in plant-water relations to understand plastic and evolutionary responses to changing climates.

@@

For plants, soil water availability is an important environmental factor in ecology and agriculture, acting as a major determinant of fitness and yield [@juenger2013natural; @greenham2017temporal]. As such, considerable effort has been placed on studying plant responses to drought, often defined conceptually and experimentally as an environmental condition of abnormally elevated aridity resulting in decreased plant performance [@passioura1996drought]. Most of the research on drought responses and tolerance strategies, including in _Brachypodium_, the focal system of this work, has involved comparisons between discrete soil water levels - control and water-limited [@des2012physiological; @edwards2012quantitative; @el2014genotype; @greenham2017temporal; @vasseur2014multivariate]. Yet soil moisture as an environmental factor is complex and multidimensional, with fluctuations varying continuously in timing, duration, and degree. Here we investigate trait responses to one important dimension of soil moisture variability - degree - with experimental and statistical approaches treating soil moisture content as a continuous variable rather than a set of fixed levels.

Quantitative genetic variation in drought resistance traits have been observed in natural populations and laboratory model systems. In particular, natural populations of Brassicaceae species (including _Arabidopsis thaliana_) and _Brachypodium_ harbor variation both constitutive and plastic traits mediating plant-water relations, including water use efficiency [@des2012physiological; @edwards2012quantitative; @greenham2017temporal; @des2017interactive], leaf chemistry [@des2017interactive; @kesari2012intron], leaf anatomy [@skirycz2011pause; @verelst2013molecular;@dittberner2018natural], root-shoot biomass partitioning [@des2012physiological; @des2017interactive], and many others [@verslues2011drought; @edwards2012quantitative; @juenger2013natural; @yarkhunova2016selection; @luo2016specific; @lenk2019transcriptional]. Among the many plant traits that can and have been measured, few have been studied more extensively than specific leaf area (SLA - often reported as its inverse, leaf mass per area or LMA). SLA provides a description of leaf architecture that is central to the leaf economics spectrum, a theory which seeks to explain variation in leaf physiological strategies, from more conservative (low SLA) to more productive (high SLA) [@wright2004worldwide]. In the context of drought stress, low SLA might be adaptive as lower leaf surface area is expected to reduce water loss through transpiration. In several cases, reductions in SLA have been reported under drought conditions [@casper2001drought]. However, for all of these traits, the shape of plastic responses to variation in water availability remain largely unmeasured.

Less is known about how trait variation and relationships between traits change across continuous environmental gradients [but see @robinson2009impact]. Genetic variation for traits can be higher, for example, in less frequently encountered environmental conditions [@gibson2004uncovering; @schlichting2008hidden; @paaby2014cryptic]. And describing this structure is important as trait variances and covariances influence evolutionary constraints. For example, if total genetic variance in trait space changes depending on the environment, then the capacity to respond to selection will vary accordingly, with reduced responses to selection under conditions where genetic variance is lower and vice versa. Similarly, if trait covariances depend on environment [@wood2015environmental], conditions which increase trait covariances may limit evolutionary potential across a range of environments by reducing the effective axes of variation [@levins1968evolution; @lande1983measurement;  @via1985genotype; @kingsolver2003environmental; @gomulkiewicz2018variation]. These phenomena are made more complex by the possibility that the relationship between trait variances and covariances with the environment may be non-linear. Investigating the genetic architecture of multiple traits is therefore useful for understanding evolution in rapidly changing environments. Fortunately, this need has been met with the development of statistical methods to reduce the dimensionality of genetic variance covariance matrices and produce meaningful summaries describing evolutionary constraints of multiple phenotypes simultaneously [@houle1992comparing; @blows2007tale; @kirkpatrick2009patterns; @kingsolver2015genetic].

To study the shape of phenotypic responses to the environment, modeling traits as mathematical functions of continuous variables, or function-valued-traits, has proven powerful [@kirkpatrick1989quantitative; @kingsolver2001variation; @griswold2008hypothesis; @stinchcombe2012genetics; @goolsby2015phylogenetic; @gomulkiewicz2018variation]. Recent decades have seen this approach used across many organisms [@gibert1998describing; @pettay2008age; @rocha2012connecting; @mason2020learning], traits [@robinson2009impact; @stinchcombe2010across; @mcguigan2010quantitative], and components of the environment [@brommer2008exploring; @mcguigan2009condition; @pearse2019life]. For plant scientists, experimental and analytical approaches leveraging the concept of function-valued-traits provide a framework for gaining a deeper understanding of plant acclimation and evolutionary adaptation to the environment.

In this study we combined these approaches into an exploratory investigation of trait plasticity, covariance and evolutionary constraints across soil moisture gradients. We model trait responses to a continuous gradient of soil moisture for multiple genotypes of two species of the model grass genus _Brachypodium_:  annual _B. distachyon_ and perennial _B. sylvaticum_. Because the shapes of trait responses cannot necessarily be known _a priori_, we use model selection among linear and non-linear environmental predictors to estimate the response function for each trait. We then estimate genotype means at different levels of soil moisture and compute the variance-covariance parameters for all traits. Finally, we ask whether patterns of variance and covariance in drying-responsive traits in _Brachypodium_ species may lead to variation in evolutionary constraint as a function of soil moisture.

@@

_Brachypodium_ is a model genus for the genetics and genomics of C3 grasses [@brkljacic2011brachypodium]. In this work we studied natural variation between and among two species of _Brachypodium_: the annual _B. distachyon_ and the perennial _B. sylvaticum_. Both species are endemic to Eurasia, with _B. distachyon_ more prevalent in seasonally dry habitats in Southern Europe, North Africa, and the Middle East, and _B. sylvaticum_ more widely distributed throughout Eurasia [@catalan2016phylogeny] (Figure \@ref(fig:maps)). The different evolutionary histories, ecological environments, and life habits of these two species may have impacted patterns of genetic variation and evolvalbility in relation to soil moisture.

To investigate genetic variation, five genotypes of each were studied to characterize patterns of variation in plant traits across an environmental gradient: _Brachypodium distachyon_ inbred lines ABR2, Adi-10, Bd21, Bd3-1, and Koz-1 and _Brachypodium sylvaticum_ inbred lines Ain-1, Ast-1, Kry-1, Osl-1, Vel-1. For each species, these genotypes represent a range of geographical origins and phenotypic diversity (Personal observations) [@steinwand2013sylvaticum; @des2017interactive]. For example, Kry-1 is native to Krym, Ukraine while Ain-1 is native to Ain Draham, Tunisia [@steinwand2013sylvaticum]. Both species are self-compatible and each of the lines used here have been maintained as inbred lines for greater than six generations [@steinwand2013sylvaticum; @vogel2009development]; as such, experimental replicates may be considered nearly isogenic.

@@

Dry down treatments began 29 and 42 days after sowing (DAS) for _B. distachyon_ and _B. sylvaticum_, respectively. Because the harvesting was divided over five consecutive days (see section below), plants were split into five equal harvest cohorts, with each cohort containing equal numbers of each watering treatment to avoid confounding harvest day with soil moisture content. Thus, though each consecutive cohort differed in age by a single day, each experienced the dry down treatment for the same amount of time. Nevertheless, we expected the difference in age between harvest cohorts to potentially impact trait expression and we therefore included harvest day (cohort) as a covariate in subsequent models. To generate a continuous gradient of final soil moisture by the end of the dry down period, plants were split into five watering treatments, receiving approximately 0, 4, 8, 12, 16, or 20 ml of water per day for 14 days (Figure \@ref(fig:racks)). Prior to initiating the experiment pots were weighed with dry soil ($mass_{dry}$) and field capacity ($mass_{max}$). During the course of the dry down, soil moisture content was calculated during the morning of day $d$ for each pot as $mass_d/(mass_{max}-mass_{dry})$. The final soil moisture reflects the combined effects of water input and output through evaporation from the soil and evapotranspiration from the plants.

@@

To characterize phenotypic responses to our experimental soil moisture gradient, we measured a suite of developmental and physiological traits. Plants were harvested in five cohorts over five days at the end of the dry down period. Each day, half of the sampled plants were harvested for above and below ground biomass, total above ground green area, $\delta C_{13}$, N content, C content. The other half were harvested and assessed for specific leaf area (SLA) and relative water content (RWC). Above ground leaf area was estimated by laying freshly harvested plants flat between plates of clear acrylic and imaging with a Nikon 5300 digital camera at a fixed distance with a 35mm Nikkor lens. Total green pixels were counted for each image with Easy Leaf Area [@easlon2014easy] (https://github.com/heaslon/Easy-Leaf-Area) with settings shown in Figure \@ref(fig:leafarea). Above ground biomass was measured after drying leaf material overnight at 60°C and then for several weeks at room temperature. Below ground biomass was measured after washing the soil matrix from roots and drying them overnight at 60°C and then for several weeks at room temperature. Above and below ground biomass was measured after leaves and roots were dried. Leaf tissues for $\delta C_{13}$, $\delta N_{15}$, nitrogen (hereafter "N") content, and carbon (hereafter "C") content were ground to a fine powder and processed by the UC Davis Stable Isotope Facility. SLA was calculated by scanning the two youngest fully emerged leaves with a 1 $cm^2$ red square. Leaf area in $mm^2$ was calculated from these same images using Easy Leaf Area. SLA was calculated as $leaf\ area / biomass_{dry}$. These leaves were also used to calculate RWC. Prior to drying, fresh leaves were weighed ($biomass_{fresh}$) and then submerged under water in 15mL falcon tubes for several hours. They were then weighed ($biomass_{turgid})$, oven-dried overnight, and weighed again ($biomass_{dry}$). RWC was calculated as $(biomass_{fresh}-biomass_{dry})/(biomass_{turgid}-biomass_{dry})$.

@@

pretty_table <- capture.output(apa_table(
 t_model_selection_summaries_byspecies
 , caption = "Model selection. 'x' indicates parameters chosen in best minimum AIC model. H = harvest day, G = genotype, E = final soil moisture, E^2 = quadratic parameter, s(E) = 2nd degree natural spline parameter",
 digits=3,
 font_size="scriptsize",
 longtable = F,
 landscape=F,
 align="c",
 midrules=1:nrow(t_model_selection_summaries_byspecies)
))

@@

### Function-valued traits
For the purposes of modeling phenotypic responses to variation in soil moisture content, we considered soil moisture content as the final soil moisture on day 14 of the dry down period for each plant, referred to in figures as $Soil\ moisture (\%_{final})$. A major challenge in studying function-valued traits is model selection. That is, identifying the functions that best describe the curvature (or lack thereof) in the shape of phenotypic responses to environmental gradients. Quadratic and natural splines have been suggested as potential functions to model non-linearities [@meyer2005randomregression], but assuming the appropriate function is problematic. Akaike information criterion (AIC) selection based on contrasting multiple complex models offers an effective means to balance predictability with over-fitting [@griswold2008hypothesis; @bolstad2010direct; @gomulkiewicz2018variation]. Because we sampled only 5 genotypes per species [@van1995linear] we treat genotype as a fixed effect in our model, which differs from random regression used elsewhere where genotypes are treated as random effects [@nussey2007evolutionary]. Thus, we began with the full fixed-effects linear regression model for traits as described below.

$Trait = H+G+E+E^2+s(E)_{df=2}+G*E+G*E^2+G*s(E)_{df=2}$

where $H = harvest\ day, \ G = genotype,\ E = soil\ moisture,\ E^2 = quadratic\ parameter,\ s(E)_{df=2} = second\ degree\ natural\ spline\ parameter$

@@

### Genetic correlations as a function of soil moisture content
We calculated trait correlations at different levels of soil moisture to characterize how genetic correlations between traits vary as a function of soil moisture content. Predicted genotypic means for each trait were calculated at 20 levels of soil moisture content (from 30% to 100% gravimetric water content) based on the model chosen by AIC (see above). Pearson correlation coefficients between genotype means were calculated at each level of soil moisture within each species.

### Plasticity through multidimensional trait space
We quantified total plasticity through multidimensional trait space as a function of soil moisture by first scaling each trait to a mean of 0 and standard deviation of 1. We then calculated the euclidean distance matrix between genotype means at all soil moisture levels. We then estimated total plasticity, measured as euclidean distance between trait values, between consecutive soil moisture levels for each genotype. $m$ is a vector of length 20 evenly spaced soil moisture measurements from 30% to 100%:

$$m=(m_1,...,m_{20})=(30\%,...,100\%)$$

If $p_{g,i} = (p_{g,1},...,p_{g,n})$ and $p_{g,i+1} = (p_{g,i+1,1},...,p_{g,i+1,n})$ are two consecutive (in terms of soil moisture) points in $n$ dimensional trait space of scaled genotype means, total plasticity (Euclidean distance) for genotype $g$ across $n$ traits at soil moisture level $i$ is calculated as

$$\Delta T_{g,i}=\sqrt{\sum_{j=1}^{n}(p_{g,i,j}-p_{g,i+1,j})^2}$$

Finally, to ask if total plasticity differed between the two species, at each level of soil moisture, we then compared the mean $\Delta{T}$ between _B. sylvaticum_ and _B. distachyon_ by T-tests. To visualize plasticity of each genotype through multivariate trait space further, we performed a principal component analysis from the matrix of scaled genotype trait means using the _prcomp_ function in R.

@@

We calculated several statistics summarizing evolutionary constraint as described in @kirkpatrick2009patterns. First, for each species, we approximated the $G$ matrix of genetic covariances by calculating variances and covariances between mean scaled genotype trait means at different levels of soil moisture. We then calculated, using the _prcomp_ function in R, the eigenvalues of each mean standardized (trait values divided by mean) $G$ matrix, $\lambda_i$. From these we then calculated the _number of effective dimensions_, $n_{D}$, equal to the sum of the eigenvalues divided by the largest eigenvalue:

$$n_{D} = \sum_{i=1}^{n} \lambda_i/\lambda_l$$

@@

$$v_T = \sum_{i=1}^{n} \lambda_i$$

(ref:curves) **Variation in phenotypic responses to soil moisture gradient as function-valued traits.** Points and the lines connecting them indicate predicted values at different levels of gravimetric soil moisture content from 30% to 100%. _B. sylvaticum_ genotypes are colored orange and _B. distachyon_ blue. (a) `r real_trait_names[1]` (b) `r real_trait_names[2]` (c) `r real_trait_names[3]` (d) `r real_trait_names[4]` (e) `r real_trait_names[5]` (f) `r real_trait_names[6]` (g) `r real_trait_names[7]` (h) `r real_trait_names[8]` (i) `r real_trait_names[9]` (j) `r real_trait_names[10]` (k) `r real_trait_names[11]` (l) `r real_trait_names[12]`

@@

@@

Across the six watering treatments, combined with random variation in water capacity of pots (Figure \@ref(fig:pots)), the dry down period resulted in a continuous environmental gradient of final soil moisture, albeit with a higher frequency of plants near the driest extreme of soil moisture variation (Figure \@ref(fig:drydown)). This gradient provides the basis for analyzing phenotypes in relation to soil moisture treated as a continuous gradient rather than limited set of discrete factors.

The observed reduction in leaf relative water content (RWC) under the driest conditions in both _B. distachyon_ and _B. sylvaticum_ indicates that at this extreme, plants were physiologically stressed (Figure \@ref(fig:curves)A). Mean leaf RWC for plants in the 10% tail of soil water content was `r tailrwc`% which is drier than that observed in the less severe soil drying conditions of Des Marais et al. [@des2017interactive]. Additional visual observations made during the experiment such as leaf rolling, another symptom of dehydration stress, were evident in plants at the lowest water treatment by the end of the dry down period.

(ref:multi) **Plasticity through multivariate trait space.** (a) Plasticity across all traits was calculated as distance between scaled phenotype for each genotype between different levels of soil moisture. Box plots indicate species median and 25th and 75th percentiles with whiskers extending to 1.5 times the interquartile range. "*" indicates significant differences between species (t-test, $\alpha < 0.05$). _B. sylvaticum_ is colored in orange and _B. distachyon_ blue.

```{r multi, fig.cap='(ref:multi)', out.width = "\\textwidth", fig.pos = "!h"}
knitr::include_graphics("../Figures/multi_plast.pdf", dpi = 108)
```

### Non-linearity in trait responses to soil moisture is common

We evaluated the degree to which traits showed linear or non-linear responses to soil moisture using an AIC model selection approach from a full model which included quadratic and natural spline parameters relating soil moisture content to plant phenotypes. We observed non-linear components (quadratic, spline, or both) in the final models for all of the traits which included an environmental (water content) predictor (Table \@ref(tab:models), Figure \@ref(fig:curves), Table \@ref(tab:outmodels), Table \@ref(tab:modelsr2)). In _B. distachyon_, all of the traits included at least one non-linear environmental predictor. In contrast total biomass and root mass did not include an environmental predictor in selected models in _B. sylvaticum_.

When considering plasticity across multidimensional trait space, it appears that most of the variation in _B. distachyon_ is attributable to responses to low soil moisture (Figure \@ref(fig:multi), Figure \@ref(fig:pca)). In contrast, _B. sylvaticum_ was more responsive to extreme wet conditions than _B. distachyon_. We observed, particularly in _B. sylvaticum_, that phenotypes were similar between extreme dry and extreme wet soil moisture contents (Figure \@ref(fig:pca)b). This similarity may be explained by the quadratic parameters of trait functions where the curvature of trait responses can lead to similar phenotypes at both environmental extremes.
